 ---
 title: "Algae time series Expression and Cell Wall thickness correlation"
 subtitle: "projectID: u2019005"
 author: "Aman Zare & Nicolas Delhomme"
 date: "`r Sys.Date()`"
 output: 
  html_document:
   fig_width: 9
   fig_height: 6
   toc: TRUE
   number_sections: TRUE
   toc_depth: 3
   toc_float:
    collapsed: TRUE
    smooth_scroll: TRUE
   code_folding: hide
   theme: "flatly" 
   highlight: pygments
   includes:
    before_body: header.html
    after_body: footer.html
   css: style.css    
 ---

# Setup 
```{r set up}
suppressPackageStartupMessages({
  library(here)
  library(ggplot2)
  library(purrr)
  library(tidyverse)
  library(UpSetR)
})
```

# Data preparation
We start by reading the DE genes and prepare a binary matrix for overlapping genes in various comparisons.
```{r data}
#read the DE files
files <- list.files(here("data/analysis/DE"), pattern = "genes\\.csv$", full.names = TRUE)

#function to calculate the frequency matrix for upset plot
fun.mat <- function(condition){
  #list of dataframes
  df_list <- lapply(files, function(file) {
    data <- read.csv(file)
    if (!is.null(condition)) {
      rows <- eval(parse(text = condition), envir = data)
      gene_names <- data[rows, 1]
    } else {
      gene_names <- data[,1]
    }
    df <- data.frame(gene = gene_names, comparison = 1)
    colnames(df) <- c("gene", gsub(".*Time-(.*)genes\\.csv", "\\1", file))
    return(df)
  })
  names(df_list) <- gsub(".*Time-(.*)genes\\.csv", "\\1", files)
  
  mat <- reduce(df_list, full_join, by = "gene")
  mat[is.na(mat)] <- 0
  return(mat)
}


# function to plot the volcano plot
fun.volcano <- function(comparison, padj=0.001){
  file <- grep(pattern = comparison, files, value = TRUE)
  data <- read.csv(file)
  data$significance <- ifelse(data$padj < padj & abs(data$log2FoldChange) >= 1,
                              ifelse(data$log2FoldChange >= 1, "up", "down"),
                              "not significant")
  x_down <- min(data$log2FoldChange) %>% round()
  x_up <- max(data$log2FoldChange) %>% round()
  y <- -log10(min(data$padj)) %>% round()
  
  # Create a data frame for the text labels
  text_data <- data.frame(
    x = c(x_down, x_up, 0),
    y = c(y, y, y),
    label = c(paste(sum(data$significance == "down"), "\n genes"),
              paste(sum(data$significance == "up"), "\n genes"),
              paste("padj <", padj))
  )
  ggplot(data, aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(aes(color = significance), alpha = 0.5) +
    scale_color_manual(values = c("up" = "red", "down" = "blue", "not significant" = "grey")) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey") +
    geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "grey") +
    geom_text(data = text_data, aes(x = x, y = y, label = label)) +
    xlab("Log2 Fold Change") +
    ylab("-Log10 Adjusted P-value")
}

```
# Volcano plots

#### {.tabset .unlisted .unnumbered .tabset-pills}

##### 1h-vs-std {-}
```{r volcano 1h-vs-std}
fun.volcano(comparison = "1h-vs-std")
```

##### 4h-vs-1h {-}
```{r volcano 4h-vs-1h}
fun.volcano(comparison = "4h-vs-1h")
```

##### 12h-vs-4h {-}
```{r volcano 12h-vs-4h}
fun.volcano(comparison = "12h-vs-4h")
```

##### 24h-vs-12 {-}
```{r volcano 24h-vs-12}
fun.volcano(comparison = "24h-vs-12")
```

##### 72h-vs-24h {-}
```{r volcano 72h-vs-24h}
fun.volcano(comparison = "72h-vs-24h")
```

##### 120h-vs-72h {-}
```{r volcano 120h-vs-72h}
fun.volcano(comparison = "120h-vs-72h")
```


# intersections (upset plot)
First we look at all intersections
```{r upset all}
mat <- fun.mat(condition = NULL)

upset(mat, 
      sets = c("72h-vs-24h", "24h-vs-12h", "12h-vs-4h", "4h-vs-1h", "1h-vs-std"),
      keep.order = TRUE, 
      mainbar.y.label = "number of overalpping genes", 
      sets.x.label = "Number of DE genes", 
      point.size = 7, 
      shade.color = "grey50", 
      text.scale = c(1.8,1.4,1.5,1.4,1.5,1.3))

upset(mat, keep.order = TRUE, order.by = "freq", sets = c("120h-vs-72h", "72h-vs-24h", "24h-vs-12h", "12h-vs-4h", "4h-vs-1h", "1h-vs-std"))
```

We might want to drop the comparisons between 120h and 72h, order the intersections by frequency and limit their numbers to the top 20 ones. 
```{r upset most}
fun.upset <- function(mat = mat, col="black") {
  upset(mat, 
        order.by = c("freq","degree"), 
        sets = c("72h-vs-24h", "24h-vs-12h", "12h-vs-4h", "4h-vs-1h", "1h-vs-std"),
        keep.order = TRUE, 
        mainbar.y.label = "number of overalpping genes", 
        sets.x.label = "Number of DE genes", 
        point.size = 7, 
        shade.color = "grey50", 
        main.bar.color = col,
        nintersects = 20, 
        text.scale = c(1.8,1.4,1.5,1.4,1.5,1.3))
}

fun.upset(mat = mat)
```

Below we look at these selected comparisons but separately for up and down regulated genes.  
For easier evaluations, we first order the bars by frequency and then by degree. This means that we will have first the bars for individual comparisons (ordered by frequency) 
and then overlapping of two comparisons followed by the group of three comparisons. 

## Up regulated genes
```{r upset up}
mat_up <- fun.mat(condition = "data$log2FoldChange >= 1")
fun.upset(mat = mat_up, col = "red")

```


## down regulated genes
```{r upset down}
mat_dn <- fun.mat(condition = "data$log2FoldChange <= 1")
fun.upset(mat = mat_dn, col = "blue")
```

Next it would be interesting to see the relationship to the cell wall related genes 
```{r upset cell wall}
cw <- read_tsv(here("data/analysis/cell-wall/goi-vst-expression_with-geneID.tsv"))
```

